Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery

### 2025-03-07 paper/markdown/supplementary_material.Rmd
DLCO <80% predictions and DLCO values
Analysis of predictions of DLCO < 80% in observations with increasing DLCO values expressed as percentages of the patient's reference. Drop-out rates and frequency of LFT and CT abnormalities included in the manuscript text.

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 10)

@@

For machine learning modeling of lung function testing outcomes, the packages _caret_ [@Kuhn2008] and [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) were utilized, which provide wrappers around the R implementations of the random forest algorithm (_ranger_) [@Wright2017; @Breiman2001], 
neural network (_nnet_) [@Ripley2014], 
support vector machines (SVM, _kernlab_) [@Weston1998; @Karatzoglou2004], 
and gradient boosted machines (GBM, _gbm_) [@Greenwell2022; @Friedman2002; @Friedman2001]. 
Model diagnostic statistics and performance evaluation metrics were computed with the development package [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra). 
SHAP (Shapley additive explanations) as a metric of global variance importance in machine learning models was calculated with the package _kernelshap_ [@Covert2020; @Lundberg2017; @Mayer2023].

@@

@@

flextable::flextable(suppl_tabs$cohort_ct) %>% 
  set_widths(c(2.3, 2.7, rep(3.9, 4))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('any CT findings: any CT findings diagnosed by a human pathologist; GGO: ground glass opacity; CTSS: human-determined CT severity score; reticulation, consolidation, bronchiectasis: diagnosed by a human radiologist; opacity and high opacity, AI: opacity and high opacity determined by artificial intelligence.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(suppl_tabs$cohort_lft) %>% 
  set_widths(c(2.3, 2.7, rep(3.7, 4))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('LFT: lung function testing; FVC: forced vital capacity; FEV1: forced expiratory volume in one second; DLCO: diffusion capacity for carbon monoxide.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  theme_vanilla

@@

flextable::flextable(suppl_tabs$cohort_symptoms) %>% 
  set_widths(c(2.3, 3.5, rep(3.7, 4))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('mMRC: modified Medical Research Council scale; ECOG: Eastern Cooperative Oncology Group performance scale.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  theme_vanilla

@@

>>>>

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 9)

@@

For machine learning modeling of lung function testing outcomes, the packages _caret_ [@Kuhn2008] and [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) were utilized, which provide wrappers around the R implementations of the random forest algorithm (_ranger_) [@Wright2017; @Breiman2001], 
neural network (_nnet_) [@Ripley2014], 
support vector machines (SVM, _kernlab_) [@Weston1998; @Karatzoglou2004], 
and gradient boosted machines (GBM, _gbm_) [@Greenwell2022; @Friedman2002; @Friedman2001]. 
Model diagnostic statistics and performance evaluation metrics were computed with the development package [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra). 
SHAP (Shapley additive explanations) as a metric of global variance importance in machine learning models was calculated with the package _kernelshap_ [@Covert2020; @Lundberg2017; @Mayer2023]. 
Moving averages of statistics of performance of machine learning models were computed with _zoo_ [@Zeileis2022].

@@

Performance of the machine learning models at prediction of reduced DLCO < 80% of the patient's reference was assessed in observations with increasing values of DLCO expressed as percentage of the patient's reference. 
To this end, we sorted the observations by increasing DLCO values and computed moving averages of accuracy and squared distance to 0/1-coded DLCO < 80% outcome. 
The moving averages were computed with the sliding window of k = 20 observations (function `rollmean()` package _zoo_). 
The squared distance to the outcome were calculated with the following formula:

$$SqDist = (p - o)^2$$

where $p$ stands for probability of DLCO < 80%, and $o$ equals to 1 for DLCO < 80% and 0 for DLCO $\geq$ 80%.

@@

flextable::flextable(suppl_tabs$cohort_ct) %>% 
  set_widths(c(2, 2.4, rep(3.2, 5))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('any CT findings: any CT findings diagnosed by a human pathologist; GGO: ground glass opacity; CTSS: human-determined CT severity score; reticulation, consolidation, bronchiectasis: diagnosed by a human radiologist; opacity and high opacity, AI: opacity and high opacity determined by artificial intelligence.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(suppl_tabs$cohort_lft) %>% 
  set_widths(c(2, 2.4, rep(3.2, 5))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('LFT: lung function testing; FVC: forced vital capacity; FEV1: forced expiratory volume in one second; DLCO: diffusion capacity for carbon monoxide.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  theme_vanilla

@@

flextable::flextable(suppl_tabs$cohort_symptoms) %>% 
  set_widths(c(2, 2.4, rep(3.2, 5))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('mMRC: modified Medical Research Council scale; ECOG: Eastern Cooperative Oncology Group performance scale.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  theme_vanilla

@@

```{r fig-dlco-class, fig.width = figur::convert(suppl_figs$dlco_class, to = 'in')$w, fig.height = figur::convert(suppl_figs$dlco_class, to = 'in')$h, fig.cap = "Accuracy of prediction of insufficient diffusion capacity for carbon monoxide by the GBM model as a function of diffusion capacity for carbon monoxide expressed as percentage of patient's reference."}

suppl_figs$dlco_class$plot

__Supplementary Figure S\@ref(fig:fig-dlco-class). Accuracy of prediction of insufficient diffusion capacity for carbon monoxide by the GBM model as a function of diffusion capacity for carbon monoxide expressed as percentage of patient's reference.__

_Predictions made by the best performing model of reduced diffusion capacity for carbon monoxide (DLCO < 80% of the patient's reference) were investigated in observations with increasing DLCO values expressed as percentage of the patient's reference._

_(A) The actual DLCO values expressed as percentages of the patient's reference are displayed in violin plots. Single observations are depicted as points, point color codes for the predicted reduced DLCO <80% in the training data and 10-repeats 10-fold cross-validation (CV). The dashed line represents the 80% cutoff used for definition of reduced DLCO,_

_(B) Moving averages of accuracy (fraction of correct predictions) at prediction of DLCO < 80% and squared distances to the 0/1-coded DLCO < 80% in the training data set and 10-repeats 10-fold CV. The moving averages calculated with the sliding window of 20 observations are displayed as solid lines. The dashed lines represent the 80% cutoff used for definition of reduced DLCO._
